---
title: "Data"
description: "Browse and download the datasets behind our reports."
page-layout: full
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(DT)
library(duckdb)
library(DBI)
library(scales)
library(yaml)

con  <- dbConnect(duckdb(),
          here::here("../cabarrus-data-pipeline/db/cabarrus.duckdb"),
          read_only = TRUE)
pop      <- dbReadTable(con, "nc_population_growth")
housing  <- dbReadTable(con, "cabarrus_housing")
landcover <- dbReadTable(con, "nlcd_landcover_area")
dbDisconnect(con, shutdown = TRUE)

dt_buttons <- function(filename, title) {
  list(
    list(extend = "csv",   filename = filename, text = "Download CSV"),
    list(extend = "excel", filename = filename, title = title,
         text = "Download Excel")
  )
}

make_dt <- function(df, filename, title, ...) {
  datatable(
    df,
    extensions = "Buttons",
    filter     = "top",
    rownames   = FALSE,
    options    = list(
      dom        = "Bfrtip",
      buttons    = dt_buttons(filename, title),
      pageLength = 25,
      scrollX    = TRUE
    ),
    ...
  )
}

codebook_html <- function(rel_path) {
  cb   <- yaml::read_yaml(here::here(rel_path))
  rows <- map_chr(cb$columns, function(x) {
    paste0(
      "<tr>",
      "<td><code>", x$name, "</code></td>",
      "<td style='white-space:nowrap'>", x$type, "</td>",
      "<td>", x$description, "</td>",
      "</tr>"
    )
  })
  cat(paste0(
    '<details class="codebook">',
    '<summary>Codebook: ', cb$title,
      ' (', length(cb$columns), ' columns)</summary>',
    '<p class="codebook-desc">', cb$description, '</p>',
    '<p class="codebook-desc"><strong>Source:</strong> ', cb$source, '</p>',
    '<table class="table table-sm table-bordered">',
    '<thead><tr>',
    '<th>Column</th><th>Type</th><th>Description</th>',
    '</tr></thead><tbody>',
    paste(rows, collapse = "\n"),
    '</tbody></table></details>'
  ))
}
```

Each dataset is derived from public sources and powers a specific report. Use the column search boxes to filter, click any header to sort, and use the export buttons to download.

::: {.panel-tabset}

## Population Growth {#pop-growth}

**Source:** U.S. Census Bureau, Decennial Census 2000/2010/2020 and Population Estimates Program (PEP) Vintage 2023.
**Used in:** [Topic: Population Growth](/topics/population/) | [Data Brief #1](/data-briefs/0001-population-growth.pdf)

::: {.panel-tabset}

### County Growth Rates {#pop-counties}

Growth rates, decade-by-decade breakdowns, and hot spot classifications for all 100 NC counties.

```{r pop-counties}
pop |>
  arrange(rank_growth) |>
  mutate(
    pct_growth  = round(pct_growth,  1),
    pct_00_10   = round(pct_00_10,   1),
    pct_10_20   = round(pct_10_20,   1),
    pct_20_now  = round(pct_20_now,  1),
    gi_star     = round(gi_star,     3),
    abs_growth  = comma(abs_growth,  accuracy = 1),
    pop_2000    = comma(pop_2000,    accuracy = 1),
    pop_2010    = comma(pop_2010,    accuracy = 1),
    pop_2020    = comma(pop_2020,    accuracy = 1),
    pop_recent  = comma(pop_recent,  accuracy = 1)
  ) |>
  select(
    County        = county_name,
    `Pop. 2000`   = pop_2000,
    `Pop. 2010`   = pop_2010,
    `Pop. 2020`   = pop_2020,
    `Pop. 2023`   = pop_recent,
    `Change`      = abs_growth,
    `% Growth`    = pct_growth,
    `% 00-10`     = pct_00_10,
    `% 10-20`     = pct_10_20,
    `% 20-Now`    = pct_20_now,
    `Rank`        = rank_growth,
    `Gi* Score`   = gi_star,
    `Hot Spot`    = spot_class
  ) |>
  make_dt("nc_population_growth", "NC Population Growth")
```

```{r pop-codebook}
#| results: asis
codebook_html("data/codebooks/nc_population_growth.yml")
```

### Top 10 Fastest-Growing {#pop-top10}

The ten fastest-growing NC counties by percentage change, 2000-2023.

```{r pop-top10}
pop |>
  slice_min(rank_growth, n = 10) |>
  arrange(rank_growth) |>
  mutate(
    pct_growth = round(pct_growth, 1),
    abs_growth = comma(abs_growth, accuracy = 1),
    pop_2000   = comma(pop_2000,   accuracy = 1),
    pop_recent = comma(pop_recent, accuracy = 1)
  ) |>
  select(
    Rank         = rank_growth,
    County       = county_name,
    `Pop. 2000`  = pop_2000,
    `Pop. 2023`  = pop_recent,
    `Change`     = abs_growth,
    `% Growth`   = pct_growth
  ) |>
  make_dt("nc_top10_growth", "NC Top 10 Fastest-Growing Counties")
```

:::

## Housing Affordability {#housing}

**Source:** U.S. Census Bureau, American Community Survey (ACS) 1-Year Estimates, 2006-2023 (no 2020 release).
**Used in:** [Topic: Housing Affordability](/topics/housing/) | [Data Brief #2](/data-briefs/0002-home-prices-wages.pdf)

::: {.panel-tabset}

### Nominal Values {#housing-nominal}

Raw dollar figures for home values, rents, and household income by year.

```{r housing-nominal}
housing |>
  select(year, home_value, gross_rent, annual_rent, med_income,
         home_to_income, rent_to_income) |>
  mutate(
    home_value     = dollar(home_value,  accuracy = 1000),
    gross_rent     = dollar(gross_rent,  accuracy = 10),
    annual_rent    = dollar(annual_rent, accuracy = 100),
    med_income     = dollar(med_income,  accuracy = 1000),
    home_to_income = round(home_to_income, 2),
    rent_to_income = round(rent_to_income, 2)
  ) |>
  select(
    Year                = year,
    `Median Home Value` = home_value,
    `Monthly Rent`      = gross_rent,
    `Annual Rent`       = annual_rent,
    `Median Income`     = med_income,
    `Price / Income`    = home_to_income,
    `Rent / Income`     = rent_to_income
  ) |>
  make_dt("cabarrus_housing_nominal", "Cabarrus Housing Nominal Values")
```

**Notes:** All values are nominal and not adjusted for inflation. Price/Income above 3.0 is conventionally considered unaffordable; Rent/Income above 0.30 indicates rent burden.

```{r housing-codebook}
#| results: asis
codebook_html("data/codebooks/cabarrus_housing.yml")
```

### Indexed (2006 = 100) {#housing-indexed}

All series indexed to 2006 to show relative growth rates.

```{r housing-indexed}
housing |>
  select(year, idx_home, idx_rent, idx_income) |>
  mutate(across(starts_with("idx_"), \(x) round(x, 1))) |>
  select(
    Year           = year,
    `Home Value`   = idx_home,
    `Annual Rent`  = idx_rent,
    `Income`       = idx_income
  ) |>
  make_dt("cabarrus_housing_indexed", "Cabarrus Housing Index (2006=100)")
```

**Notes:** Values above 100 indicate growth relative to 2006.

:::

## Land Cover {#land-cover}

**Source:** USGS/MRLC National Land Cover Database (NLCD), 30-meter resolution, years 2011, 2016, 2021.
**Used in:** [Topic: Vanishing Farmland](/topics/farmland/) | [Data Brief #3](/data-briefs/0003-vanishing-farmland.pdf)

### Acreage by Class and Year {#landcover-area}

Estimated acres of each land cover class within Cabarrus County.

```{r landcover-area}
landcover |>
  mutate(acres = round(acres, 1)) |>
  pivot_wider(names_from = year, values_from = acres,
              names_prefix = "") |>
  arrange(class) |>
  select(
    `Land Cover Class` = class,
    `2011`             = `2011`,
    `2016`             = `2016`,
    `2021`             = `2021`
  ) |>
  make_dt("cabarrus_nlcd_landcover", "Cabarrus County NLCD Land Cover (acres)")
```

**Notes:** Pixel counts multiplied by 900 m² per cell and converted to acres (1 acre = 4,046.86 m²). The NLCD uses consistent multi-date methodology designed for temporal comparison; small year-to-year fluctuations reflect classification uncertainty, not actual land change.

```{r landcover-codebook}
#| results: asis
codebook_html("data/codebooks/nlcd_landcover_area.yml")
```

:::

```{=html}
<script>
document.addEventListener("DOMContentLoaded", function () {
  const hash = window.location.hash;
  if (!hash) return;
  const target = document.querySelector('[data-bs-target="' + hash + '"]');
  if (target) {
    bootstrap.Tab.getOrCreateInstance(target).show();
    setTimeout(function () {
      document.querySelector(hash)?.scrollIntoView({ behavior: "smooth" });
    }, 150);
  }
});
</script>
```
